\chapter{Conclusion}\label{ch:conclusion}

\section{Summary \& Findings}

This thesis addressed the problem of automatically reconstructing
scenario-specific Interaction Models from \Cpp\ codebases and execution
traces in event-driven systems.
The core challenge is that static analysis alone produces over-approximate,
scenario-agnostic graphs, while trace-based analysis alone produces
under-complete models due to limited runtime instrumentation.
No prior approach simultaneously addresses publish-subscribe topology
recovery and scenario-scoped querying of the call graph, nor targets \sysml
as the output formalism.

The proposed hybrid approach builds a static call graph in \neo (including
IDC pub-sub topology-resolved edges) and, separately, collects timestamped
execution traces into runtime logs. The two evidence sources are kept
separate: they are combined only as a set union of their interaction edge
sets when the reconstructed models are compared during evaluation.
Scenario-scoped Cypher queries then narrow the static call graph (and,
equivalently, the trace-derived sequence table) to the interactions
relevant to a single scenario.
The scenario-scoped sequence table is projected into \sysml textual
notation.

The approach was developed and evaluated on \emph{CPSCore}, an open-source
event-driven \Cpp\ framework, across four scenarios: three
happy-path scenarios occurring naturally in CPSCore's existing code,
plus one (S4) purpose-built to isolate an evidence-combination precondition the other
three cannot exercise.
Three hypotheses were tested to answer the research questions:

\begin{description}
  \item[H1 (confirmed).]
    Combining static and dynamic evidence produces more accurate interaction models than any single
    evidence source.
    The full pipeline (C5) achieves the highest macro-F1 of any condition
    tested (0.964) with perfect recall on every scenario.
    On the three happy-path scenarios, the static graph is already
    reachability-complete, so combining is redundant with static analysis
    alone (C4$=$C2). On the constructed scenario~S4, where a
    publish-subscribe boundary and log suppression each hide a different
    interaction, combining is strictly additive: F1(C4)$=$1.000, versus
    0.667 for either single source.
    This confirms that the combination's value activates precisely when no single
    source can see the full picture, and is otherwise safely redundant
    (\cref{sec:h1-results}).

  \item[H2 (confirmed).]
    Code-graph-guided instrumentation eliminates all measured noise
    (50\% fewer retained nodes, 70\% fewer edges) without losing coverage.
    A blinded six-rater expert panel confirmed perceptual superiority:
    guided output was rated strictly higher in 47 of 54 comparisons, with
    one reversal traceable to a specific design trade-off (collapsing
    repeated property-read calls into a single edge).
    The quantitative and perceptual results align: noise removal improves
    usability (\cref{sec:h2-results,sec:h2-qual}).

  \item[H3 (confirmed).]
    In a holistic five-condition experiment: the reconstructed diagram
    alone achieves perfect component recall, interaction recall and rated
    completeness on every scenario. It matches or exceedes all combinations
    that include source code and traces.
    Conditions without the diagram consistently miss a suppressed-but-source-visible
    interaction that the diagram states explicitly.
    The diagram is therefore the distinguishing factor, not merely a
    supplement (\cref{sec:h3-holistic}).
\end{description}

As an exploratory complement to H1--H3, a practitioner hands-on session and
interview study with four engineers on a second, much larger industrial
codebase (\cref{sec:practitioner}) provides encouraging evidence that the
\emph{static half} of the approach holds at industrial scale. This session
exercised only the static-extraction and graph-querying capability. It did not
involve any runtime tracing or dynamic analysis.
The key scenario exercise involved a proxy-leak defect whose root cause is an
implicit event-driven interaction invisible without reconstructing the call
chain. It was resolved using only the static graph. This mirrors H1's
finding that static analysis preserves fault-path recall where traces
cannot.
The industrial system's architecture (proxy-based, virtual interfaces,
dynamic service binding) is the kind of system where H1 predicts C4$>$C2
would materialise (\cref{sec:cpscore-scale,sec:h1-c4c2}). Because the
session collected no dynamic evidence, this remains a motivated prediction
rather than a measured result, and is left to future work.

A single cross-cutting pattern unifies all four evidence sources
(\cref{sec:discussion}): \textbf{additional evidence helps exactly when
the primary evidence source is missing something and is otherwise
redundant.}
Combining evidence helps on fault paths but not happy paths (H1). Diagrams help comprehension 
when traces are incomplete but not when they are already sufficient (H3). Guided 
instrumentation improves usability by removing what is irrelevant (H2).
Furthermore, practitioners solved a fault-path defect using the exact evidence 
source (static graph) that H1 identifies as the recall-preserving complement 
to incomplete traces. This pattern also generates a concrete, falsifiable 
prediction for future work: on systems with cross-module virtual dispatch, combining
should show C4$>$C2 on \emph{all} scenario types, not just fault
paths (\cref{sec:future-work}).

\section{Practical Implications}

The pipeline enables an engineer working on an event-driven \Cpp\ system to
obtain a scenario-specific \sysml\ interaction diagram by issuing
natural-language queries against the property graph, rather than spending
several working days manually tracing call chains, cross-referencing header
files and correlating log output.

On CPSCore, the static extraction skill populated 2,574 ordered interactions
across 121 components in a single Cypher query execution. A further Cypher
query identified the cross-component call boundaries to trace (nine caller
functions across five files); the \texttt{add-runtime-instrumentation} skill
then instrumented the corresponding call-site patterns with
\texttt{csp\_matcher}, requiring no manual call-graph reading.
Because the pipeline is codebase-agnostic (skills query the live schema, not
hardcoded labels), switching to a new target requires only re-running the
static extractor and re-invoking the same skills.
The practitioner session confirmed this transfer cost is low in practice:
four engineers on a much larger industrial codebase used the tool
productively within a single session and the structural and dependency
diagrams were rated consistently useful for architectural comprehension.

\section{Contributions}

The primary contributions of this thesis are organised into three research
contributions and three engineering artefacts:

\paragraph{Research contributions.}
\begin{enumerate}
  \item \textbf{A static property graph and a separate trace pipeline}
    whose interaction edge sets are compared while remembering which
    evidence each interaction came from (static, dynamic, or both).
    Keeping the two sources separate but comparable is what lets the
    evaluation report exactly where each edge originated
    (\cref{sec:pg-schema}).

  \item \textbf{A scenario-scoping method} that trims the call graph down
    to just the interactions relevant to one scenario using plain
    scenario-scoped Cypher queries, without re-running the analysis from
    scratch (\cref{sec:combine-scope}).

  \item \textbf{A clear, evidence-based account} of when combining static
    and dynamic analysis actually helps and when it does not. Combining
    protects fault paths and adds new interactions at indirection
    boundaries (H1). Guided instrumentation removes noise without losing
    coverage (H2). Finally, the reconstructed diagrams help comprehension
    specifically when the trace is incomplete (H3).
\end{enumerate}

\paragraph{Engineering artefacts.}
\begin{enumerate}\setcounter{enumi}{3}
  \item A \textbf{small, reversible instrumentation tool}
    (\texttt{csp\_matcher}) that inserts and removes trace probes
    automatically, driven by a simple find/replace specification
    (\cref{sec:csp-matcher}).

  \item A \textbf{set of eleven natural-language agent skills} that run the
    whole pipeline, from source code to \sysml\ diagram on any codebase
    without code changes (\cref{sec:skill-design}).

  \item A \textbf{way to recover publish-subscribe wiring} that ordinary
    call-graph analysis cannot see, by matching subscription and
    publication call sites that share the same channel identifier
    (\cref{sec:macro-resolution}).
\end{enumerate}

\section{Limitations \& Threats to Validity}

The most significant limitation is scale: all three hypotheses were
evaluated on a single open-source system (CPSCore, $\approx$13\,kLOC)
across four scenarios.
CPSCore's absence of cross-module virtual dispatch means the static call
graph is already reachability-complete, explaining why combining is
redundant on happy-path scenarios. In larger systems with virtual dispatch
or dynamic plugin loading, the combination benefit would likely be more
pronounced.
The practitioner session on an industrial-scale codebase
(\cref{sec:practitioner}) provides encouraging qualitative evidence of
transfer, but does not substitute for a controlled replication.

The H3 experiment used the same LLM family both to generate the C5 diagram
and to carry out the comprehension task that judged it. Because the same
model plays both roles, the two are not fully independent (a
generator--consumer circularity).
The H2 six-rater panel preserved effect direction on every dimension,
but a larger sample would strengthen confidence in magnitude.
Ground-truth diagrams are inherently subjective, and the evaluation
metric does not capture temporal ordering accuracy.

A complete discussion of threats to validity, including the researcher
degrees of freedom (the analysis choices that could have biased the result)
identified and corrected during H1/H2, is given in
\cref{ch:analysis}.

\section{Future Work}\label{sec:future-work}

Several directions are open for future investigation:

\begin{description}
  \item[Controlled evaluation on a second system.]
    The practitioner session (\cref{sec:practitioner}) suggests the
    static approach transfers to larger, industrial codebases, but only
    qualitatively. Extending the fault-path/dual-gap scenario set beyond
    the single constructed scenario (S4) currently reported would be a
    natural first step. Applying the full, pre-registered
    H1/H2/H3 hypothesis tests to that codebase or to other event-driven
    \Cpp\ systems (for example, using different publish-subscribe frameworks such as
    ROS~2 or MQTT), would turn this encouraging signal into a controlled
    generalisability result.

  \item[Diagram labelling and interactivity.]
    The practitioner interviews (\cref{sec:interview-findings})
    independently converged on two concrete requests. The first was
    richer, less abstract event/message labels on sequence diagrams
    (\eg inline source class and line number on hover). The second was
    interactive exploration (search, filter, collapse/expand) rather than
    static, one-shot diagrams. Both are natural extensions of the
    source-location data the pipeline already carries (\cref{sec:utility}).

  \item[Temporal ordering accuracy.]
    The current evaluation abstracts over message ordering.
    Developing a metric that accounts for the relative order of interactions
    in the generated diagram would provide a more complete picture.

  \item[Incremental updates.]
    As the codebase evolves, the \ac{PG} should be updatable incrementally
    rather than re-built from scratch.
    Neo4j's support for live graph updates makes this tractable. But, the
    scenario-scoping invalidation logic would need to be designed carefully.

  \item[LLM-assisted diagram naming.]
    Function names in the generated diagrams are currently taken verbatim
    from the source code.
    A post-processing step using a \ac{LLM} could produce more
    human-readable message labels, improving the practical utility of the
    output.

  \item[Integration with Renaissance DB.]
    The \ac{GDI} project maintains an architecture model database
    (Renaissance DB) that could serve as a richer source of module-level
    metadata, potentially reducing the number of unresolved black-box
    components.
\end{description}

\section{Reproducibility}

The complete source code, experiment data, skill definitions, and scoring
scripts used in this thesis are available at:

\begin{center}
\url{https://github.com/prathikanand7/masters-thesis-work.git}
\url{https://ci.tno.nl/gitlab/anand_krishnan.prathik-tno/cpscore.git}
\end{center}
