\chapter{Evaluation}\label{ch:analysis}

This chapter evaluates the \ac{GDI} pipeline against the three research
questions introduced in \cref{ch:introduction} by testing the three
hypotheses (H1, H2, H3) that operationalise them.
It reports the design and results of each experiment on the CPSCore case
study, together with a complementary practitioner validation session on a
second industrial codebase.
It also reflects on the engineering utility of the approach and discusses
threats to validity, including several methodological risks that were
identified and corrected over the course of the experiments themselves.

\section{Evaluation Design}\label{sec:eval-design}

The evaluation is organised around the three hypotheses rather than around
the research questions directly, since each hypothesis is what makes a
research question testable:

\begin{itemize}
  \item \textbf{H1} (\cref{sec:h1}) tests \rqtwo\ by comparing reconstruction
    accuracy across the five conditions C1--C5 (\cref{sec:conditions}) on
    four scenarios (\cref{sec:scenarios}): three occurring naturally in
    CPSCore's existing codebase (S1--S3), plus one (S4) purpose-built to
    test a specific fusion precondition that S1--S3 cannot exercise.
  \item \textbf{H2} (\cref{sec:h2}) tests \rqthree\ by comparing guided
    against full instrumentation (\cref{sec:guided-vs-full}) on structural
    compression, noise, and coverage, complemented by a blinded expert
    rating of the resulting diagrams.
  \item \textbf{H3} (\cref{sec:h3}) is a downstream, task-based test:
    whether the reconstructed models improve an agent's performance on a
    concrete scenario-comprehension task, relative to reasoning over raw
    source code.
\end{itemize}

\rqone\ is answered directly by the property graph schema design
(\cref{sec:pg-schema}) rather than by an experiment, and is addressed
separately in \cref{sec:rq-answers}.
\Cref{sec:practitioner} reports a fourth, exploratory data point
(practitioner feedback from a second, much larger industrial codebase)
that is deliberately kept separate from H1--H3 since it was not run as a
controlled or pre-registered comparison.

\section{H1: Does Fusion Improve Reconstruction Accuracy?}\label{sec:h1}

\textbf{Hypothesis.} Combining static analysis and dynamic instrumentation
(C4) improves interaction model reconstruction accuracy over any single
source (C1--C3) alone.

\subsection{Metric}

For every condition and every scenario, the generated \sysml sequence
diagram is compared against a manually authored reference diagram over
\emph{interaction edges} (unique ordered $\langle\text{sender},
\text{receiver}\rangle$ pairs) using precision, recall, and F1:

\[
  P = \frac{|\text{gen} \cap \text{ref}|}{|\text{gen}|}, \quad
  R = \frac{|\text{gen} \cap \text{ref}|}{|\text{ref}|}, \quad
  F_1 = \frac{2PR}{P + R}
\]

where $\text{gen}$ is the set of interaction edges produced under a given
condition and $\text{ref}$ is the corresponding reference edge set for that
scenario. \newline
Here, $P$ denotes precision, $R$ denotes recall, and $F_1$ denotes the
harmonic mean of precision and recall.
Reference diagrams were authored manually per scenario, before any condition
was scored against them, to avoid tuning the reference to a particular
condition's output.

This precision/recall/F1 formulation follows the standard
information-retrieval convention~\cite{manning:ir} and mirrors its recent
use for comparing statically-computed against dynamically-sampled call
graphs~\cite{helm:totalrecall}.
It is also the established convention within the sequence-diagram
reconstruction literature itself, where F-score and recall are used to
score a reconstructed diagram against a reference~\cite{taniguchi:seqtrace}.
Per-scenario F1 scores are then aggregated as described next
(\cref{sec:h1-corrections}).

\subsection{Methodological Corrections}\label{sec:h1-corrections}

Four issues in the experimental design were identified and corrected while
H1 was being run, rather than after the fact:

\begin{description}
  \item[Ground-truth circularity.] An early version of the reference
    diagrams was partly derived from the same static extraction used to
    produce the C2/C4/C5 conditions, which would have inflated their scores
    by construction. The reference diagrams were reauthored independently of
    any single condition's output.
  \item[\texttt{SCENARIO\_PRIMARY} scope boundary.] The boundary used to
    decide which functions count as ``in scope'' for a scenario was
    initially defined \emph{after} inspecting the results of a preferred
    condition, risking a scope definition that favoured that condition.
    The scope boundary (the \texttt{SCENARIO\_PRIMARY} criterion) is now
    fixed from the scenario definition alone, before any condition is run.
  \item[F1 aggregation ambiguity.] It was initially unclear whether F1
    should be computed per-scenario and averaged (macro-F1) or computed
    once over edges pooled across all scenarios (micro-F1). The two give
    materially different answers when scenario sizes are unequal.
    Macro-F1 was adopted (mean of per-scenario F1 scores, weighting
    each scenario equally) because it prevents large scenarios from
    dominating the aggregate~\cite{sokolova:metrics}.
    Micro-F1 is also reported. Values agree to within $\pm 0.013$.
  \item[Constructed-scenario disclosure.] S4 was added after S1--S3 were
    scored, specifically to test whether fusion could ever exceed \emph{both}
    single-source conditions at once. This is a property no naturally-occurring
    CPSCore code path exhibits (\cref{sec:scenarios}). Reporting S4 blended
    into the same aggregate as S1--S3 without comment would risk implying the
    effect is typical of the codebase rather than a deliberately engineered
    edge case. S4 is therefore reported as its own stratum
    (\cref{tab:h1-results}) and discussed separately
    (\cref{sec:h1-c4c3}), never folded silently into the S1--S3 numbers.
\end{description}

\subsection{Results}\label{sec:h1-results}

\Cref{tab:h1-results} reports precision, recall, and F1 for all five
conditions, stratified by scenario type, over the reported four-scenario
set (S1--S4).

\begin{table}[htb]
\centering
\caption{H1 per-scenario results: precision / recall / F1 by condition.
  S1--S3 occur naturally in CPSCore; S4 is constructed (\cref{sec:scenarios}).}
\label{tab:h1-per-scenario}
\small
\begin{tabular}{@{} l c c c c @{}}
\toprule
\textbf{Condition} & \textbf{S1} & \textbf{S2} & \textbf{S3} & \textbf{S4} \\
\midrule
C1: LLM-only        & 0/0/0.00    & 0/0/0.00    & 0/0/0.00    & .50/.50/.50 \\
C2: Static-only     & 1.00/1.00/1.00 & 1.00/1.00/1.00 & .60/1.00/.75 & 1.00/.50/.67 \\
C3: Dynamic-only    & 1.00/1.00/1.00 & 1.00/1.00/1.00 & 1.00/1.00/1.00 & 1.00/.50/.67 \\
C4: Static+Dynamic  & 1.00/1.00/1.00 & 1.00/1.00/1.00 & .60/1.00/.75 & 1.00/1.00/\textbf{1.00} \\
C5: Full pipeline & 1.00/1.00/1.00 & 1.00/1.00/1.00 & .75/1.00/.86 & 1.00/1.00/\textbf{1.00} \\
\bottomrule
\end{tabular}
\end{table}

\Cref{tab:h1-results} reports the same data aggregated by scenario type,
now decomposed into precision, recall, and F1 (macro-averages). S4 is kept
as its own column rather than folded into the S1--S3 stratum, since it
is the only constructed (not found-in-the-wild) scenario in the suite
(\cref{sec:scenarios}).

\begin{table}[htb]
\centering
\caption{H1 stratified results: macro-averaged precision, recall, and F1 by
  condition and scenario type (reported four-scenario set: S1--S4).}
\label{tab:h1-results}
\small
\begin{tabular}{@{} l ccc ccc ccc @{}}
\toprule
& \multicolumn{3}{c}{\textbf{Happy-path (S1--S3)}}
& \multicolumn{3}{c}{\textbf{Constructed (S4)}}
& \multicolumn{3}{c}{\textbf{Overall (S1--S4)}} \\
\cmidrule(lr){2-4} \cmidrule(lr){5-7} \cmidrule(lr){8-10}
\textbf{Condition} & P & R & F1 & P & R & F1 & P & R & F1 \\
\midrule
C1: LLM-only       & 0.000 & 0.000 & 0.000
                   & 0.500 & 0.500 & 0.500
                   & 0.125 & 0.125 & 0.125 \\
C2: Static-only    & 0.867 & 1.000 & 0.917
                   & 1.000 & 0.500 & 0.667
                   & 0.900 & 0.875 & 0.854 \\
C3: Dynamic-only   & \textbf{1.000} & \textbf{1.000} & \textbf{1.000}
                   & 1.000 & 0.500 & 0.667
                   & \textbf{1.000} & 0.875 & 0.917 \\
C4: Static+Dynamic & 0.867 & 1.000 & 0.917
                   & 1.000 & 1.000 & 1.000
                   & 0.900 & \textbf{1.000} & 0.938 \\
C5: Full pipeline & 0.917 & \textbf{1.000} & 0.952
                   & 1.000 & \textbf{1.000} & 1.000
                   & 0.938 & \textbf{1.000} & \textbf{0.964} \\
\bottomrule
\end{tabular}
\end{table}

F1 aggregation: macro-average (mean of per-scenario F1 scores).
Micro-F1 values agree to within $\pm 0.013$ (see \cref{sec:h1-corrections}).

\subsection{Discussion: Why C4 $=$ C2 on S1--S3}\label{sec:h1-c4c2}

The central finding on the three scenarios that occur naturally in CPSCore's
existing codebase is that \textbf{C2 (static-only) and C4 (static+dynamic)
produce numerically identical results on all three found-in-the-wild
scenarios (S1--S3)}.
This is not a null result nor a pipeline error. It follows directly from a
structural property of CPSCore described in
\cref{sec:cpscore-scale}: the codebase contains no cross-module virtual
dispatch, so every call target is statically deterministic.
Consequently, every interaction edge the trace discovers was already present
in the static call graph, making the dynamic evidence a strict subset of the
static evidence, \emph{on S1--S3 specifically}:

\[
  \text{C3} \subseteq \text{C2} \quad \text{(on S1--S3)}
  \;\Longrightarrow\;
  \text{C4} = \text{C2} \cup \text{C3} = \text{C2} \quad \text{(on S1--S3)}
\]

In a system with cross-module virtual dispatch, dynamic plugin loading, or
reflection-based routing, traces could observe targets that static analysis
cannot resolve. In such a system C4 would strictly exceed C2.
CPSCore's existing, found-in-the-wild code does not exhibit this property on
S1--S3, and the C4$=$C2 identity there is the expected mathematical
consequence. \Cref{sec:h1-c4c3} shows what happens once a boundary with this
property is deliberately constructed (S4).

\paragraph{C3 (dynamic-only) is the best happy-path condition, not a
redundant subset of the static graph.}
On the reported four-scenario set, C3 achieves macro-P/R/F1 of
1.000/1.000/1.000 on S1--S3, strictly above C2/C4's 0.867/1.000/0.917.
C2/C4 both \emph{carry} S3's two static
over-approximation false positives (\texttt{flush}, \texttt{instance})
that C3 never sees at all, since those calls simply do not execute during
the S3 test. On S1--S3, fusion is therefore not
protective (there is no suppressed evidence to protect against) and
is mildly \emph{harmful} to precision relative to the trace alone. The
entire empirical case for fusion's value rests on S4
(\cref{sec:h1-c4c3}), the sole scenario purpose-built to require it.

\paragraph{Precision deficit of C2/C4.}
Decomposing beyond F1 reveals a second finding hidden by the aggregate:
C2 and C4 both have macro-precision~$=$~0.867 on the happy-path stratum, not
1.000.
The deficit comes from S3, where the static graph includes two
Synchronization$\to$Logging calls (\texttt{flush} and
\texttt{instance}) that are real source-level call sites but fall outside
the S3 reference scope (they belong to
\texttt{SynchronizedRunner.runSynchronized}, not to the S3 primary function
\texttt{SimpleRunner.runStage}).
Static analysis cannot scope to a single scenario without slicing. This is
precisely the over-approximation problem that motivates C5.

\subsection{Discussion: S4 and the Limits of C4 $=$ C2}\label{sec:h1-c4c3}

S1--S3 all satisfy $\text{C3} \subseteq \text{C2}$
(\cref{sec:h1-c4c2}), which places a hard ceiling on fusion: C4 can match C2
but never exceed it there. S4 (\cref{sec:scenarios}) was constructed
specifically to remove that ceiling on \emph{both} sides at once, by
introducing a component boundary, namely a \texttt{boost::signals2} publish/
subscribe bridge between \texttt{StageEventBridge} (Synchronization) and
\texttt{StageEventListener} (Aggregation), where each single-source
condition is blind to a different edge:

\begin{itemize}
  \item C2 sees \texttt{stream} (the fault-path log call is a normal,
    statically-resolvable function call) but \textbf{not}
    \texttt{onStageEvent}, because the signal-to-slot dispatch is
    established via \texttt{connect()} at runtime and has no resolvable call
    expression in the static graph: P/R/F1~$=$~1.000/0.500/0.667.
  \item C3 sees \texttt{onStageEvent} (the slot fires and is traced) but
    \textbf{not} \texttt{stream}, because a \texttt{LogLevel::NONE}
    suppression idiom makes the \texttt{CPSLOG\_ERROR} call a no-op at
    runtime: P/R/F1~$=$~1.000/0.500/0.667.
  \item C4, the union of both, recovers both edges:
    P/R/F1~$=$~1.000/1.000/\textbf{1.000}.
\end{itemize}

\textbf{F1(C4)~$=$~1.000 strictly exceeds both F1(C2)~$=$~0.667 and
F1(C3)~$=$~0.667 on S4.} This is the empirical confirmation of the
hypothesis that motivated constructing S4: fusion can be strictly
\emph{additive}, not merely protective, whenever static and dynamic
analysis are blind to different edges rather than one being a subset of
the other. All five S4 conditions were executed for real, not simulated
(see \cref{sec:scenarios} for the provenance of each).

\paragraph{C5 (provenance-aware) inherits this result.} C5 gives the
agent C2 and C3 as separately labelled, provenance-tagged evidence
(\cref{sec:h1-c5}), which lets it retain both \texttt{onStageEvent} and
\texttt{stream} on S4, reaching F1~$=$~1.000, matching C4 exactly.
\Cref{sec:h1-c5} reports this methodology and its results across the full
four-scenario set.

\subsection{Discussion: C5 Provenance-Aware Synthesis}\label{sec:h1-c5}

C5 gives the agent C2 and C3 as two \emph{separately labelled} evidence
sources, each edge tagged with whether the other source also observed
it, together with explicit guidance to reason about \emph{why} a
single-source edge might be missing from the other (genuine runtime
suppression, an indirect dispatch invisible to static analysis, or a
genuine static over-approximation) rather than treating single-source
status itself as weak evidence. The agent's response also includes a
complete, ordered call sequence (the real dynamic trace order as backbone,
with retained static-only edges inserted at the position implied by the
scenario description), giving the reconstructed sequence diagram a
temporal structure a flat, unordered element list would not carry.

\paragraph{Result: recall parity with C4, at a small, localised precision
cost.} With provenance-tagged evidence, C5 retains
\texttt{CPSLogger::instance} on S1 and \texttt{stream} on
S4, matching C4's recall (R~$=$~1.000) on every one of the four
scenarios. The only imprecision appears on S3, where the same
``a suppressed single-source edge is usually real'' reasoning that
correctly recovers S4's fault-path edge is applied, incorrectly, to
\texttt{flush}: a real Synchronization$\to$Logging call, but one that
originates from a different function entirely outside S3's defined scope.
The agent has no way to distinguish this from a genuine fault-path
suppression using only component-level provenance. C2's element list
records source, target, and interaction name, but not the calling
function, so ``suppressed-but-in-scope'' and ``real-but-out-of-scope'' are
indistinguishable inputs to the same reasoning step. On the full
four-scenario set (S1--S4), C5 achieves macro-P/R/F1 of
0.938/1.000/\textbf{0.964}. This is the highest F1 of any of the five conditions,
including C4's raw union (0.938) and C3's raw trace (0.917, held back by
its S4 recall miss).

\paragraph{Value chain summary.}
The full value chain, on the four-scenario set, is: fusion is
redundant with static analysis on S1--S3 (\cref{sec:h1-c4c2}) and is
strictly additive only on the constructed S4 case
(C4~$>$~C2 \emph{and} C4~$>$~C3). Provenance-aware agent synthesis then
\emph{preserves} that recall (C5~R~$=$~C4~R~$=$~1.000 on every
scenario) while also partially correcting C4's precision where static
analysis over-approximates (S3), for the best overall reconstruction
accuracy of any condition tested, though not the best precision, since C3
has none at all in this set. Agent synthesis with provenance-aware
prompting is therefore a net improvement over the raw fused union, not a
trade against it. However, the disclosed S3 imprecision shows that a
component-level (rather than function-level) provenance granularity still
leaves one scope-disambiguation gap unresolved.

\section{H2: Does Guided Instrumentation Improve Scenario Focus?}\label{sec:h2}

\textbf{Hypothesis.} Code-graph-guided instrumentation (\cref{sec:guided-vs-full})
improves scenario focus by suppressing irrelevant execution detail, relative
to full, unguided instrumentation.

H2 has a quantitative arm (structural compression and noise/coverage
metrics, complete) and a qualitative arm (blinded expert usability rating,
complete).
Both were required before the hypothesis could be called confirmed: H2 is
fundamentally a claim about human-perceived usability. The quantitative
metrics alone cannot establish that a smaller, less noisy graph is actually
easier for an engineer to use (\cref{sec:h2-corrections}).

\subsection{Metrics}

Let $G_{\text{full}}$ and $G_{\text{guided}}$ denote the property-graph
subsets produced by full and code-graph-guided instrumentation respectively
for a given scenario, and let $R$ denote the scenario-relevant reference
node set for that scenario (the same reference used for H1).
Four metrics are computed over nodes (and separately over edges):

\[
  \text{NodeReduction} = 1 - \frac{|V(G_{\text{guided}})|}{|V(G_{\text{full}})|},
  \qquad
  \text{EdgeReduction} = 1 - \frac{|E(G_{\text{guided}})|}{|E(G_{\text{full}})|}
\]

\[
  \text{Noise}(G) = \frac{|V(G) \setminus R|}{|V(G)|},
  \qquad
  \text{Coverage}(G) = \frac{|V(G) \cap R|}{|R|}
\]

Here, $V(G)$ and $E(G)$ denote the node set and edge set of graph $G$
respectively, $R$ is the scenario-relevant reference node set,
$G_{\text{full}}$ is the property graph produced by full (unguided)
instrumentation, and $G_{\text{guided}}$ is the property graph produced by
code-graph-guided instrumentation.
$\text{Noise}(G)$ is the proportion of nodes retained by a configuration that
are \emph{not} scenario-relevant (lower is better). $\text{Coverage}(G)$ is
the proportion of scenario-relevant nodes that are actually retained (higher
is better, with $1.000$ meaning no relevant node was lost).
Noise and Coverage are the retained-set analogues of $1-\text{Precision}$ and
Recall~\cite{manning:ir}.
The noise concept itself is adapted from its use in software architecture
recovery evaluation, where the proportion of irrelevant retained elements is
likewise termed noise~\cite{constantinou:noise}. Here it is applied to
instrumentation scoping rather than clustering-based architecture recovery,
so the mapping is an adaptation of that concept rather than a direct
transfer from either source.
A configuration that reduces noise by discarding relevant nodes along with
irrelevant ones is not actually an improvement. Coverage is what rules that
failure mode out.

\subsection{Circularity Check}\label{sec:h2-corrections}

Guided instrumentation restricts probes to functions inside the static
dependence slice for the scenario, and the reference set $R$ used to score
noise and coverage is itself derived from the same scenario scope boundary
(\cref{sec:h1-corrections}).
This means a noise score of exactly $0$ for $G_{\text{guided}}$ is partly
guaranteed \emph{by construction}: if the slice and the reference share a
scope definition, everything the slice keeps is relevant almost by
definition.
This was treated as a real risk rather than dismissed. Two things address
it:
\begin{itemize}
  \item The result is explicitly reported as \emph{partly by construction}
    rather than presented as a clean empirical finding.
  \item Scenario S3 served as a genuine falsification test: unlike the other
    scenarios, its slice and reference boundaries were not defined from the
    same pass. A noise score of $0$ on S3 is therefore empirical evidence,
    not a tautology. S3 did in fact reproduce the pattern seen elsewhere, which is
    what allows the result to be reported as more than circular.
\end{itemize}

\subsection{Quantitative Results}\label{sec:h2-results}

\begin{table}[htb]
\centering
\caption{H2 per-scenario results: full vs.\ guided instrumentation scored
  against H1 reference diagrams (TP/FP/FN over interaction edges). S4 is
  constructed and reported separately (\cref{sec:h2-s4}).}
\label{tab:h2-per-scenario}
\small
\begin{tabular}{@{} l l c c c c c c @{}}
\toprule
\textbf{Scenario} & \textbf{Variant} & \textbf{Interactions} & \textbf{TP} & \textbf{FP} & \textbf{FN} & \textbf{Coverage} & \textbf{Noise} \\
\midrule
S1 & Full    & 9 & 4 & 5 & 0 & 1.000 & 0.556 \\
S1 & Guided  & 4 & 4 & 0 & 0 & 1.000 & 0.000 \\
\addlinespace
S2 & Full    & 9 & 1 & 8 & 0 & 1.000 & 0.889 \\
S2 & Guided  & 1 & 1 & 0 & 0 & 1.000 & 0.000 \\
\addlinespace
S3 & Full    & 9 & 3 & 6 & 0 & 1.000 & 0.667 \\
S3 & Guided  & 3 & 3 & 0 & 0 & 1.000 & 0.000 \\
\addlinespace
S4 & Full    & 10 & 2 & 8 & 0 & 1.000$^\dagger$ & 0.800 \\
S4 & Guided  & 1 & 1 & 0 & 1 & 0.500 & 0.000 \\
\bottomrule
\end{tabular}
\end{table}

$^\dagger$ S4-full's coverage is a scoring artifact, not genuine recovery of
the fault-path edge; see \cref{sec:h2-s4}.

\Cref{tab:h2-results} summarises the mean metrics across all three scenarios.

\begin{table}[htb]
\centering
\caption{H2 aggregate results: guided vs.\ full instrumentation (mean across
  S1--S3).}
\label{tab:h2-results}
\begin{tabular}{@{} l c c c @{}}
\toprule
\textbf{Metric} & \textbf{Full} & \textbf{Guided} & \textbf{Reduction} \\
\midrule
Nodes (lifelines)     & 6.0 & 3.0 & $-50\%$ \\
Interaction edges     & 9.0 & 2.7 & $-70\%$ \\
Coverage              & 1.000 & 1.000 & $0\%$ \\
Noise                 & 0.704 & 0.000 & $-100\%$ \\
\bottomrule
\end{tabular}
\end{table}

Guided instrumentation reduced the retained node count by \textbf{50\%}
and interaction edges by \textbf{70\%} relative to full instrumentation,
while driving noise to \textbf{0} and holding coverage at \textbf{1.000}:
every scenario-relevant interaction was retained despite the aggressive
reduction in diagram size.
Full-trace false positives are real runtime events from non-primary
components active in the same test run (\eg
\texttt{Aggregation$\to$Logging:stream},
\texttt{Configuration$\to$Logging:stream}) that are architecturally valid
calls but irrelevant to the scenario under analysis.

\paragraph{Why S4 is excluded from this comparison.}
The full-vs-guided contrast only has something to measure when the trace
actually contains extra, off-scenario events for guided instrumentation to
filter out.
In S4, the suppression that defines its fault-path edge
(\texttt{CPSLogger::instance()->setLogLevel(LogLevel::NONE)},
\cref{sec:scenarios}) acts at the logging level, not at the instrumentation
level: the relevant call site is never logged regardless of whether the
probe scope is full or guided.
Full and guided instrumentation therefore produce the \emph{same} (empty or
partial) trace on S4, so a noise/coverage comparison between them would
report a vacuous $0\%$ difference that says nothing about the guided
mechanism. It would only be restating the fault-path suppression already
characterised in H1.
The H2 quantitative arm is consequently scoped to the three scenarios
(S1--S3) where full instrumentation actually produces excess, off-scenario
trace events for guided instrumentation to remove.

\subsection{S4: A Discovered Scoring Artifact}\label{sec:h2-s4}

S4 (\cref{sec:scenarios}) was extended into H2 after being added to H1, for
the same reason: it is the one scenario where static and dynamic evidence
each miss a \emph{different} edge. Unlike S1--S3, whose
\texttt{full\_trace.txt} is a direct copy of a single pre-existing capture
that already contains those scenarios' code paths, S4's production code did
not exist when that capture was taken. Its full-instrumentation trace was
therefore constructed honestly as "the pre-existing capture plus S4's own
test-captured row" — the best available simulation of what a full
instrumentation pass across the now-larger test suite would produce.

Scoring this file produced an unexpected result: \textbf{S4-full shows
coverage~$=$~1.000, appearing to recover \texttt{stream} as well as
\texttt{onStageEvent}.} Tracing the matched row shows this is not
genuine. The pre-existing capture already contains several
unrelated \texttt{Synchronization$\to$Logging:stream} rows, produced by
other functions (\eg \texttt{SimpleRunner.runStage},
\texttt{SynchronizedRunnerMaster.runStage}) under normal (non-suppressed)
logging conditions during unrelated test runs, none of them
\texttt{StageEventBridge::publishStage}. Because the scoring convention
matches only on the bare $(\text{source}, \text{target},
\text{interaction-name})$ triple (\cref{sec:h1-corrections}), deliberately,
to mirror how a human or agent describes an interaction, any one of those
unrelated rows satisfies the S4 reference's
\texttt{Synchronization$\to$Logging:stream} requirement. The genuine S4
fault-path call remains absent from every trace, full or guided: it is
suppressed by the same \texttt{LogLevel::NONE} runtime gate discussed
above, which is orthogonal to instrumentation scope. The guided trace
cannot exhibit this artifact by construction: it only ever contains rows
captured during the S4 test itself, and correctly reports coverage~$=$~0.500
(1 of 2 reference edges).

\textbf{Why this strengthens, rather than weakens, the H2 argument.} It
reveals a failure mode of full instrumentation beyond noise: pooling trace
events across an entire, unscoped test suite increases the chance that some
unrelated call elsewhere in the codebase coincidentally satisfies a
reference triple by bare name alone, producing \emph{unverifiable, inflated
coverage} rather than an unambiguous true positive. Guided instrumentation,
scoped by construction to only the events the scenario's own execution
produces, cannot exhibit this failure mode. S4-full's raw coverage number is
retained in \cref{tab:h2-per-scenario} for transparency
(\cref{sec:h1-corrections}). Neither instrumentation strategy recovers
\texttt{stream} for S4.
S4 is therefore excluded from the S1--S3 aggregate in
\cref{tab:h2-results}, matching how it is kept as a separate stratum
throughout H1 (\cref{sec:h1-corrections}).

\subsection{Qualitative Arm: Blinded Expert Rating}\label{sec:h2-qual}

Structural compression is necessary but not sufficient to claim that guided
instrumentation produces a \emph{more usable} diagram. That is a claim
about human perception, not about graph size.
A blinded A/B rating protocol was designed to test this directly, and was
extended from an initial 2-rater round to a 6-rater panel to strengthen
confidence in the direction and stability of the effect:

\begin{itemize}
  \item Six external expert raters, each a domain engineer with direct
    knowledge of the codebase, independently rated diagrams generated
    under full and guided instrumentation without being told which
    configuration produced which diagram.
  \item A rating form and a private decode key (held separately from the
    raters) were used to keep the assignment blind during scoring. The
    diagram labelled \emph{A} versus \emph{B} was re-randomised per
    scenario so that the label itself carried no information about the
    underlying condition.
  \item All six raters returned completed forms scoring S1--S3 on three
    5-point Likert dimensions (accuracy, readability, and practical
    usefulness) for both conditions, together with free-text comments and
    an overall preference judgement.
\end{itemize}

\paragraph{Results.}
\Cref{tab:h2-qual-results} reports the mean rating (across all six raters)
per scenario and dimension.

\begin{table}[htb]
\centering
\caption{H2 qualitative results: mean blinded expert ratings (1--5 scale,
  $n=6$ raters) for full vs.\ guided instrumentation, per scenario and
  rating dimension.}
\label{tab:h2-qual-results}
\begin{tabular}{@{} l cc cc cc @{}}
\toprule
 & \multicolumn{2}{c}{\textbf{Accuracy}} & \multicolumn{2}{c}{\textbf{Readability}} & \multicolumn{2}{c}{\textbf{Usefulness}} \\
\textbf{Scenario} & Full & Guided & Full & Guided & Full & Guided \\
\midrule
S1 & 2.33 & 4.67 & 2.33 & 4.83 & 3.17 & 4.50 \\
S2 & 2.33 & 4.17 & 2.33 & 4.83 & 2.67 & 2.83 \\
S3 & 2.50 & 5.00 & 2.67 & 5.00 & 3.00 & 4.67 \\
\midrule
\textbf{Mean} & \textbf{2.39} & \textbf{4.61} & \textbf{2.44} & \textbf{4.89} & \textbf{2.94} & \textbf{4.00} \\
\bottomrule
\end{tabular}
\end{table}

Guided instrumentation was never rated \emph{below} full instrumentation on
accuracy or readability by any rater on any scenario: 34 of the 36
accuracy/readability comparisons show guided strictly preferred, and the
remaining 2 (both rater R4, on S2 and S3 accuracy) are exact ties.
Usefulness was rated higher or tied for guided in all but one of the 18
rater--scenario usefulness judgements. Across all $6 \times 9 = 54$
rater--scenario--dimension comparisons, guided instrumentation was rated
strictly higher in \textbf{47}, tied in \textbf{6}, and rated strictly
lower in exactly \textbf{1}.
Averaged across all scenarios and dimensions, guided instrumentation
scored \textbf{4.50} against \textbf{2.59} for full instrumentation.

Expanding the panel from two to six raters did not change the direction of
the effect, but it surfaced heterogeneity a two-rater sample could not.
On accuracy and readability the effect is unanimous in direction: no rater
rated full instrumentation higher on any scenario (twice tied, never
reversed). Usefulness is unanimous for guided on S1 and S3 but genuinely
split on S2, the one scenario where guided instrumentation collapses seven
repeated \texttt{Configuration}$\to$\texttt{Logging} property-read calls
into a single edge (\cref{tab:h2-per-scenario}): three raters tied it, one
rated full instrumentation higher, and two preferred guided. The common
thread in the free-text comments is that collapsing the repeated reads
discards a \emph{count} some raters consider useful for debugging, even
where it did not affect their accuracy or readability scores. The full
per-rater decomposition, the individual free-text comments, and the
raters' improvement suggestions are reported in
\cref{sec:appendix-h2-perrater}.

With all six raters' ratings now in hand, \textbf{H2 is confirmed, and
more robustly so than the original two-rater sample suggested}: guided
instrumentation not only compresses the graph without losing
scenario-relevant nodes (quantitative arm, \cref{sec:h2-results}), it is
also rated more accurate and more readable by every one of six external
domain experts in a blinded comparison, on every scenario (qualitative
arm, above). The one qualification is usefulness on S2, where the
node/edge reduction discards a property-read count that multiple raters
independently considered relevant to debugging.

\section{H3: Do Reconstructed Models Improve Task Performance?}\label{sec:h3}

\textbf{Hypothesis.} Agents equipped with reconstructed structural and
sequence diagrams perform software engineering tasks more accurately and
efficiently than agents relying only on source code and execution traces.

H3 was scoped to a single downstream task (Scenario Comprehension)
rather than the full space of tasks an agent might perform with an
interaction model, in order to keep the evaluation tractable within the
project timeline.
A single holistic experiment was conducted, with \textbf{five conditions}
that isolate source code, execution traces, and the reconstructed C5
diagram both individually and in combination, run across the same four
scenarios (S1--S4).

\subsection{Protocol}

Ground truth was derived directly from the H1 reference diagrams: the
same \texttt{elements.json} files (manually source-read, independent of
any pipeline condition) were converted into a scoring schema of
expected components, expected interactions, and a reference explanation.

For each scenario, an LLM agent was prompted to identify the
participating components, enumerate the cross-component interactions,
and explain the scenario execution flow, under one of five conditions:

\begin{description}
  \item[Source only.] Agent receives relevant source snippets (function
    bodies) for the scenario. No trace, no diagram.
  \item[Trace only.] Agent receives the runtime trace slice for the
    scenario. No source, no diagram.
  \item[Diagram only.] Agent receives the C5 interaction list (fused
    static+dynamic evidence, agent-synthesised). No source, no trace.
  \item[Source + Trace.] Agent receives both the source snippets and the
    runtime trace slice.
  \item[Source + Trace + Diagram.] Agent receives the source snippets,
    the runtime trace slice, and the C5 interaction list.
\end{description}

Three of the five conditions (trace only, source+trace, and
source+trace+diagram) reuse real agent responses collected against this
same harness and ground truth in earlier isolated comparisons. The
remaining two (source only, diagram only) were newly collected for this
holistic design, so that every point in the design space (each context
source alone, and the combinations leading up to full context) is
populated with real, non-fabricated data from the same model and prompt
family.

Component recall and interaction recall apply the same recall formulation
used for H1 (\cref{sec:h1}) to two element types: nodes and edges
respectively.
Completeness is scored on a 1--5 rubric.
For the three reused conditions, completeness was rated blind (responses
anonymised, decode key held separately), following standard blinding
practice for reducing rater bias in software engineering
experiments~\cite{wohlin:experimentation}. The two newly collected
conditions (source only, diagram only) were rated in a single, non-blind
pass by the same rater applying the same rubric. This is disclosed
explicitly as a methodological limitation (\cref{sec:threats-rdof}).

\subsection{Holistic Comparison: Source, Traces, and Diagram in Isolation and Combination}
\label{sec:h3-holistic}

\begin{table}[htb]
\centering
\caption{H3 holistic comparison: five conditions across the reported
  scenario set (S1--S4).}
\label{tab:h3-holistic}
\small
\begin{tabular}{@{} l c c c @{}}
\toprule
\textbf{Condition} & \textbf{Component recall} & \textbf{Interaction recall} & \textbf{Completeness (1--5)} \\
\midrule
Source only               & 0.917 & 0.800 & 4.25 \\
Trace only                 & 0.917 & 0.900 & 4.25 \\
Diagram only                & \textbf{1.000} & \textbf{1.000} & \textbf{5.00} \\
Source + Trace              & 0.917 & 0.900 & 4.25 \\
Source + Trace + Diagram    & \textbf{1.000} & \textbf{1.000} & \textbf{5.00} \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Diagram only and source+trace+diagram are the standout results of
this experiment, and they are identical on every measure.} With
\emph{no} access to source code and \emph{no} access to the runtime
trace (only the C5 interaction list), diagram only achieves perfect
component recall (12/12), perfect interaction recall (10/10), and perfect
completeness (5.00/5.00) on every scenario, including S4, the constructed
dual-gap scenario. Source+trace+diagram matches it exactly on all three
measures. This is a genuinely strong result in favour of H3: a
reconstructed diagram alone is not merely a helpful supplement, it can be
a \emph{sufficient} basis for scenario comprehension. Adding
source code and traces on top of it neither helps nor hurts, because the
diagram alone already saturates every metric in this scenario set.

Source only has the weakest interaction recall of all five
conditions (0.800). The agent's narrative under-reports macro-based,
log-call interactions (\texttt{RAIILogStream.stream}) on S1 and S4, even
though the source snippets provided contain explicit comments
identifying them. This is not a lack of access. The information is
present in the prompt. Rather, it is a tendency for source-grounded narration to
foreground the ``main'' control-flow path (the trace-confirmed,
listener-connected path) and omit secondary, log-only branches. Trace
only avoids this failure mode on S1 (it directly observes the log
emission event when it fires) but repeats it on S4, where the trace
itself never captures the suppressed edge (\texttt{LogLevel::NONE}).
Diagram only avoids it everywhere, because C5 (\cref{sec:h1-c5}) surfaces
the interaction as an explicit, labelled edge regardless of whether it
was exercised at runtime or requires source-level inference to notice.

Source+trace and source+trace+diagram reproduce the same
pattern as the single-source conditions. Source+trace scores
identically to trace only in aggregate (0.917/0.900). Adding source
snippets to the trace does not fix the missed \texttt{stream} edge on
S4, for the same source-narration reason described above. This is a
more specific failure than simply ``having both sources does not help'':
inspecting the source+trace response directly shows the agent implicitly
treats the trace as the more authoritative source whenever the two
disagree on what happened, rather than treating source and trace as two
independent, equally weighted evidence streams to reconcile. On S4, the
source snippet is annotated with an explicit comment identifying the
\texttt{CPSLOG\_ERROR} call as ``source-visible (static analysis sees
this call unconditionally)'' and separately noting it is
``suppressed at runtime.'' However, the agent's response narrates only the
trace-confirmed, listener-connected path and never surfaces the
suppressed branch as a distinct interaction at all. Having source access
in addition to the trace does not correct this. Once the trace supplies
a concrete, single ``here is what happened'' account, the agent appears
to anchor on it, downweighting the source-only branch as background
detail rather than treating it as an independent fact to reconcile against
the trace.
Source+trace+diagram recovers to perfect recall (1.000/1.000). The
diagram avoids the same failure mode not by \emph{outweighing} the trace
with a competing narrative, but by removing the asymmetry that produces
it in the first place: C5's interaction list presents every edge as a
flat, equally weighted fact
(\texttt{Source}~$\to$~\texttt{Target}~:~\texttt{Interaction}) with no
``this is what actually happened at runtime'' framing for the model to
treat as more authoritative than anything else. There is consequently
nothing for the agent to implicitly rank above the suppressed edge, and
no comment-driven false assumption for it to build a hallucinated
priority ordering on top of. This confirms that it is specifically the
\emph{diagram} (not source, not traces, not their combination) that
reliably supplies the missing edge across every condition it appears in,
and suggests the mechanism is a de-biasing one: the diagram guides the
model with flat, unranked facts rather than implicitly signalling that
one evidence source (the trace) should be trusted over another (the
source) when the two are read together.

Completeness mirrors the recall pattern exactly. The two
conditions that include the diagram (diagram only and
source+trace+diagram) both achieve perfect completeness (5.00) on
every scenario, including S4. Source only, trace only, and source+trace
are tied at 4.25, all three losing a point specifically on S4, the
same scenario where they also lose interaction recall, for the same
underlying reason (the missed \texttt{stream} edge).

\subsection{H3 Conclusion}

\textbf{H3 is supported, with the strongest and most direct evidence
coming from the two conditions that include the C5 diagram.} Across all
four scenarios, both happy-path (S1--S3) and the constructed dual-gap
scenario (S4), diagram only and source+trace+diagram achieve
identical, perfect scores on every metric (component recall, interaction
recall, and completeness), matching or outperforming every other
condition, including combinations that additionally include source code
and/or execution traces. These are the only two conditions that achieve
perfect recall on every scenario. The three conditions without the
diagram (source only, trace only, source+trace) are all held to
0.917/0.800--0.900 recall by the same recurring failure: under-reporting
the log-suppression interaction that the diagram states explicitly.

This is a more precise, and more strongly supported, conclusion than a
design that isolates the diagram's benefit to a single scenario would
provide: here the diagram's benefit is visible not only on S4 but as the
distinguishing factor across the \emph{entire} condition-by-condition
comparison, and diagram only's standalone perfect performance (matched,
not exceeded, by adding source and traces on top of it) shows that the
reconstructed model is not merely a supplement to source and traces but
can substitute for both.

This mirrors the H1 finding (\cref{sec:h1-c5}): provenance-aware agent
synthesis (C5) is what allows the reconstructed model to retain the
evidence that makes it useful downstream. The holistic result is a
genuine, independent cross-validation, obtained from a second measurement
instrument (a structured comprehension task rather than an edge-set F1
score), that the same mechanism responsible for H1's improved
reconstruction accuracy also improves a consuming agent's task
performance. In short, agents equipped with the reconstructed C5 diagram,
whether alone or alongside source code and traces, consistently recover
the log-suppression interaction that source- and trace-based conditions
miss, with equal or greater completeness on every scenario tested. The
benefit is not one of \emph{access}, since the source-only and
source+trace conditions have the same information available and still fail
to surface it, but of presentation: the diagram's flat, unranked facts
counteract the agent's tendency to narrate around the trace-confirmed
path at the expense of source-only, not-exercised evidence. Limitations
(one constructed scenario, two non-blind conditions) are discussed in
\cref{sec:threats-rdof}.

\section{Practitioner Validation: A Second Industrial Codebase}\label{sec:practitioner}

The H1--H3 experiments validate the approach in depth on a single system
(CPSCore) under controlled, pre-registered conditions.
This section reports a complementary, exploratory data point.
\emph{Renaissance MCP}, the tool built in this project (and the wider
\ac{GDI} effort), exposes the Renaissance static-extraction backend and
graph-querying approach as an MCP integration inside GitHub Copilot.
It was trialled by practitioners at an industrial partner on a second
\Cpp\ codebase several orders of magnitude larger than CPSCore.

\textbf{This is not a repeat of H1/H2/H3.}
It is an unblinded, semi-structured hands-on session and interview study
with a convenience sample of four engineers, not a controlled comparison
against a reference diagram.
It is reported here because it speaks directly to the external-validity gap
identified elsewhere in this chapter (\cref{sec:threats-rdof}): whether the
underlying approach holds up on a second, much larger industrial system.
Codebase and defect details below are kept deliberately generic, to avoid
disclosing any partner-internal implementation specifics.

\subsection{Session Design}

Four engineers participated in a structured hands-on session, followed by a
semi-structured interview.
Each session had four parts: (i) tool installation and setup; (ii) an
\emph{estimation game} in which the participant first guessed, then used
the tool to check, several quantitative facts about the codebase;
(iii) a \emph{scenario exercise} in which the participant used the tool to
investigate a real, previously-resolved engineering issue; and (iv) a
wrap-up discussion, including questions about the tool's various diagram
outputs.

\Cref{tab:renaissance-scale} reports the six estimation-game answers,
illustrating the scale of the second codebase relative to CPSCore
(\cref{tab:impl-summary}: 2,574 static interactions, 121 components).

\begin{table}[htb]
\centering
\caption{Estimation-game answers, illustrating codebase scale on the
  second industrial system.}
\label{tab:renaissance-scale}
\begin{tabular}{@{} l r @{}}
\toprule
\textbf{Query} & \textbf{Answer} \\
\midrule
Method declarations on the largest class (a test helper) & 923 \\
Header files not included elsewhere (excl.\ externals)   & 9,606 \\
Inclusions of the most-included header file              & 2,053 \\
Calls to the most-called function                         & 23,846 \\
Events declared by a representative event-heavy interface & 118 \\
Call sites of a representative header-declared method      & 0 (unused) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Scenario Exercise: A Real Engineering Issue}\label{sec:scenario-exercise}

The scenario exercise was built around a previously-resolved defect in the
codebase, chosen because its root cause is exactly the class of problem
this thesis targets: an implicit, event-driven interaction that is
invisible without reconstructing the call/event chain.

\textbf{Defect summary (generalised and anonymised).}
An internal control service exposes several interfaces to a client
component via proxy objects, created on connect.
On service disconnect, most proxies are correctly freed. One proxy was
not due to a missing reset in the connection-handling code.
After a later service restart, the stale proxy silently stopped
delivering change-notification events, so a downstream component was no
longer updated on a subsequent event.
This defect was visible only by tracing the event path from the service
reconnect handler, through the stale proxy, to the downstream event
handler.

\textbf{Task.}
Participants were asked to use the tool to answer three questions about the
\emph{already-fixed} version of the codebase: which files were changed to
fix the issue, which test exercises the fix, and where the affected proxy
objects are used in the code and what they do.
All three questions are answerable from the static call/containment graph
alone (no runtime trace is needed to find where a proxy is constructed,
used, and released). This makes this a check on the same static-extraction
and query capability underlying \rqone\ and \rqtwo, applied to an
unfamiliar, industrial-scale codebase rather than to CPSCore.
Participants successfully identified the modified connection-handling file,
the relevant crash-recovery unit test, and the proxy's construction/release
sites: the same components underlying the actual historical fix.

\subsection{Interview Findings}\label{sec:interview-findings}

Four practitioners were interviewed: a software designer working on a core
subsystem (P1), a software product owner for an embedded control
service (P2), a senior, long-tenured software architect responsible for
inter-component synchronisation (P3), and an algorithm engineer
(P4).
Interviews covered general tool usefulness and specific feedback on six
diagram types, grouped into three categories: class structure (class
diagram, polymetric view), object interaction (call graph, dependency
diagram), and system overview (structural diagram, sequence diagram).

\paragraph{General usefulness.}
All four participants reported using AI coding assistants daily and saw
value in architecture/dependency information for their work.
P3 reported completing an investigation task in a five-minute tool session
that had previously taken roughly five months.
This is a striking self-report, but it rests on a single, unverified,
retrospective comparison and is reported here as an anecdote rather than
as a measured effect.
P2 noted a specific failure mode of code-search-based assistants that the
graph-based tool avoided: locating a stub implementation whose concrete
definition lived outside the immediately visible workspace.
P4 found the tool most useful for onboarding onto unfamiliar code (\eg
after time away from a component) and for locating test and requirement
coverage gaps.

\paragraph{Diagram feedback.}
Feedback converged on a consistent pattern across participants: diagrams
showing \emph{structure and dependency} (dependency diagrams, structural
diagrams) were consistently rated useful and, in P3's case, judged cleaner
and more representative of the actual implementation than the team's
existing modelling-tool output.
Sequence diagrams were valued for surfacing race conditions, flaky tests,
and cross-component effects that participants did not expect (P1, P3, P4).
However, all three raised the same concern: event/message labels (\eg a generic
\texttt{Subscribe} event) were too abstract to interpret without additional
context, and they requested inline detail (source class, line number) on hover.
The class diagram and polymetric view received the most mixed feedback:
three of four participants preferred the plain class diagram over the
polymetric view, and the polymetric view's size-based visual encoding
(method/attribute count as width/height) was repeatedly flagged as
measuring the wrong thing: P2 and P4 both noted that visual size does
not indicate where business logic or complexity actually lives, and P4
suggested cyclomatic complexity or requirements traceability as more
informative dimensions than raw member count.
Call graphs were the most polarising output: P1 and P4 found them
either redundant with build-system-derived views they already had (P4) or
not what they expected from a call graph (P3), while P2 rated the
equivalent class-dependency diagram positively for indicating how many
``hops'' separate two components.

\paragraph{Cross-cutting improvement requests.}
Three requests recurred across multiple interviews, independently of
diagram type. First, \emph{interactivity}: static, one-shot diagrams were
repeatedly compared unfavourably to previous tools that allowed
search, filtering, and collapsing/expanding substructure (P1, P2, P3).
Second, \emph{non-code context}: several participants wanted the tool to
incorporate information the property graph does not currently model, such
as timing/logging data, deployment configuration, and machine/product
variant (P2, P4). This is a direct instance of the construct-validity
limitation already noted for this thesis's own evaluation
(\cref{sec:h1-corrections}): the graph only contains what was extracted,
and several practically important questions live outside that scope
entirely. Third, \emph{invocation friction}: more than one participant
felt they had to phrase requests unusually explicitly for the tool to
trigger at all (P1), or wanted clearer guidance on which task types the
tool is and is not suited to (P4).

\subsection{Relationship to This Thesis's Contribution}

The class diagram and polymetric view are contributed by other members of
the \ac{GDI} project and fall outside the scope of this thesis. They are
reported here only as context for the tool participants experienced as a
whole.
The findings most relevant to this thesis's own contribution are the
consistently positive reception of the dependency and structural diagrams
(directly descended from the interface dependency table,
\cref{sec:idt}) and the specific, actionable critique of the sequence
diagram's message labelling (directly relevant to the \sysml generation
stage, \cref{sec:sysml-generation}). The request for inline source/line
detail on hover is a concrete, low-effort extension of the existing
\texttt{sourceLineNumber} provenance already carried on \texttt{CppCalls}
edges (\cref{sec:utility}), and it is noted as future work
(\cref{sec:future-work}).

\paragraph{Connection to H1 and the C4$=$C2 prediction.}
The proxy-leak defect (\cref{sec:scenario-exercise}) is structurally
analogous to the constructed dual-gap scenario S4 in H1
(\cref{sec:h1-c4c3}): an implicit, event-driven interaction that is
architecturally present in source (the stale proxy's missing
\texttt{reset} in connection-handling code) but invisible at runtime
unless the specific failure condition (service disconnect followed by
reconnect) is exercised. Participants resolved it using only the static
call/containment graph, the same evidence source H1's static-only
condition (C2) provides and that recovers architecturally-present
interactions runtime tracing alone is liable to miss
(\cref{sec:h1-c4c2}). This is informal but convergent evidence that the
H1 finding generalises beyond CPSCore: \emph{static analysis recovers
fault-path interactions that traces are unlikely to exercise.}

The same architecture also bears on the C4$=$C2 prediction
(\cref{sec:cpscore-scale,sec:h1-c4c2}). Unlike CPSCore, the industrial
codebase uses proxy objects, virtual interfaces (118 events on a single
representative interface), and dynamic service-connect/disconnect
patterns, exactly the cross-module virtual dispatch where traces would
discover call targets invisible to static analysis and C4$>$C2 is
predicted to materialise. P2's observation that code-search-based
assistants fail to resolve stub implementations whose concrete
definitions live behind indirection confirms the mechanism. The
prediction was measured directly on CPSCore by S4, where a single
\texttt{boost::signals2} indirection boundary makes C4 strictly exceed
both C2 and C3 (\cref{sec:h1-c4c3}). The practitioner session and S4
therefore corroborate the same mechanism from two independent
directions: one observed in a real, much larger industrial codebase, the
other constructed and measured on CPSCore.

\section{Discussion: Cross-Cutting Pattern}\label{sec:discussion}

A single pattern emerges across all four evidence sources (H1, H2, H3,
and the practitioner session): \textbf{additional evidence helps exactly
when the primary evidence source is missing something, and is otherwise
redundant.}

\begin{description}
  \item[H1 (fusion).] On happy-path scenarios found in CPSCore's existing
    code, the static graph already contains every interaction the trace
    could provide; fusion adds nothing (C4$=$C2). On fault-path scenarios,
    where runtime logging is suppressed, the static graph fills in what the
    trace misses; fusion is protective (C4$>$C3). On the one scenario
    constructed to isolate a cross-module indirection boundary (S4), each
    single source is missing a \emph{different} interaction, and fusion
    becomes additive rather than merely protective (C4$>$C2 and C4$>$C3
    simultaneously).
  \item[H2 (slicing).] Guided instrumentation removes all noise without
    losing coverage, because the slicing criterion correctly identifies
    what the scenario needs. The six-rater expert panel confirmed
    that this translates into perceived usability, with the single
    exception (S2 usefulness reversal) occurring precisely where the removed
    information was \emph{itself} useful for a different task (debugging
    property-read counts).
  \item[H3 (comprehension).] The reconstructed diagram improves agent
    task performance only on the fault-path scenarios where the trace is
    incomplete. This is exactly where H1 shows fusion adds value. When the trace
    already covers everything (happy paths), the diagram is redundant.
  \item[Practitioner session.] The proxy-leak defect was resolved using
    only the static graph, a fault-path interaction invisible to a
    normal-execution trace. This mirrors the mechanism H1 demonstrates on
    S4: static analysis sees what traces cannot exercise.
\end{description}

The implication is that the pipeline's value is not uniform but
\emph{conditional on information gaps}: it earns its keep precisely when
a simpler approach has a blind spot. This is a practically useful finding
because it tells engineers \emph{when} to invoke the full pipeline
(fault paths, rare execution paths, event-driven indirection) versus when
a single evidence source suffices (well-behaved, fully-exercised control
flow).

The pattern also generates a falsifiable prediction for future work: on a
codebase with cross-module virtual dispatch (unlike CPSCore), fusion
should produce C4$>$C2 on \emph{all} scenario types, not just fault
paths, because the static graph will have blind spots even on
happy-path execution. The practitioner session's industrial codebase
(proxy-based, 118-event interfaces, dynamic service binding) is exactly
such a system, and the engineers' difficulty resolving stub
implementations behind indirection is evidence of precisely this
static-graph blind spot.

\section{Answering the Research Questions}\label{sec:rq-answers}

\subsection*{\rqone: Property Graph Schema}

\textit{What intermediate property graph representation can integrate
static call graphs and timestamped runtime events for interaction model
reconstruction?}

The 13-node, 13-relationship schema described in \cref{tab:node-types,tab:edge-types}
demonstrates that a typed \ac{PG} can co-locate all artefacts
required for hybrid reconstruction without loss of provenance.
Static structure is encoded in \texttt{CppCalls},
\texttt{CppInherits}, and \texttt{CppContains} edges extracted by
Renaissance.
Runtime events are encoded as additional edges tagged with wall-clock
timestamps, anchored to the same \texttt{CppFunctionDefinition} nodes
already in the graph.

The critical enabling property is that the \ac{PG} is an open,
additive representation: runtime events are ingested incrementally
without invalidating the static sub-graph, and scenario-specific
Cypher queries over the merged graph require no structural changes
to accommodate new scenarios.
The agent skill layer (\cref{sec:skill-design}) further demonstrates
that scenario scope can be expressed as a plain-language query and
translated into a Cypher traversal at runtime, satisfying the
scenario-flexibility requirement without modifying the tool code.

\subsection*{\rqtwo: Hybrid vs. Individual Sources}

\textit{How does combining static analysis and incomplete traces improve
accuracy and completeness of scenario-specific interaction model
reconstruction?}

The H1 results (\cref{sec:h1-results}) give a more precise and more honest
answer than ``yes, hybrid wins'': fusion improves reconstruction accuracy
\emph{conditionally}, not universally.
On happy-path scenarios found in CPSCore's existing code, static analysis
alone matches the full pipeline: the trace adds nothing that the static
graph did not already provide.
On fault-path scenarios, fusion measurably improves over dynamic-only
analysis, because the static graph fills in error-handling and recovery
interactions that a single trace recording is unlikely to have exercised.
A third, constructed scenario (S4, \cref{sec:h1-c4c3}) shows that fusion's
advantage is not limited to being protective: once a component boundary is
introduced where static and dynamic analysis are each blind to a
\emph{different} interaction (here, a \texttt{boost::signals2}
publish/subscribe indirection), fusion becomes strictly additive, beating
both single sources at once (F1~$=$~1.000 vs.\ 0.667 for both C2 and C3).
The practical implication is that the value of combining sources depends on
both what kind of behaviour the scenario exercises and what kind of dispatch
mechanism connects its components, and a thesis that only tested
happy-path scenarios over directly-resolvable calls would have concluded
fusion adds no value at all, which is why S4 was deliberately constructed
to isolate this precondition.

\subsection*{\rqthree: Dependence-Based Slicing}

\textit{How can dependence-based slicing filter trace events and constrain
static interactions to only scenario-relevant behaviour?}

The H2 results (\cref{sec:h2-results}) show that code-graph-guided
instrumentation (applying the slicing criterion \emph{before}
instrumentation rather than filtering after the fact) reduces retained
nodes by 50\% and edges by 70\% relative to full
instrumentation, while holding coverage of scenario-relevant behaviour at
$1.000$.
This demonstrates the filtering mechanism works structurally.
The H3 holistic results (\cref{sec:h3-holistic}) provide complementary
evidence from the consumer side: diagram-augmented context improves agent
comprehension specifically on fault-path scenarios where slicing recovers
evidence that runtime tracing suppresses.
The H2 expert rating (\cref{sec:h2-qual}) confirms that the structural
improvement also translates into improved human-perceived usability:
guided instrumentation was rated more accurate and more readable than full
instrumentation by both blinded raters across all three scenarios.

\section{Engineering Utility}\label{sec:utility}

Beyond the precision/recall metrics, the primary claim of a tool
development thesis is that the tool is useful to engineers.
Three concrete utility properties were demonstrated during development,
each traceable to a specific architectural decision in the skill-based
pipeline (\cref{sec:skill-design}).

\paragraph{Pipeline automation via composable skills.}
The full path from \Cpp\ source to a \sysml sequence diagram is
executed by chaining the twelve agent skills described in
\cref{tab:skills}.
Each skill is stateless and produces an auditable intermediate artefact:
the \texttt{get-neo4j-schema} skill emits \texttt{graph\_schema.txt},
the \texttt{get-sequence-dependency-table} skill emits a CSV, and the
\texttt{draw-sysml-sequence-model} skill emits the final \texttt{.sysml}
file.
An engineer who disagrees with a generated interaction can inspect
the CSV row, find the \texttt{sourceLineNumber} property on the
corresponding \texttt{CppCalls} edge via \texttt{cypher-to-output},
and verify or override the edge directly in the graph without
rerunning the extraction.
Because skills query the graph schema dynamically rather than
hardcoding node labels, the same skill set works on any codebase
analysed by Renaissance without modification.

\paragraph{Scale.}
The \texttt{get-sequence-dependency-table} skill extracted 2,574
ordered interactions from CPSCore in a single Cypher query execution.
Constructing an equivalent ordered interaction list manually (by
reading source files and cross-referencing header includes) would take
an experienced engineer several working days.
The \texttt{get-runtime-traces} skill required editing nine function
bodies across five files, guided entirely by a mapping CSV generated
from a single Cypher query against the graph. No knowledge of the
full codebase structure was required to identify the instrumentation
targets.

\paragraph{Correctness preservation.}
Inserting \texttt{TRACE\_FUNCTION\_SCOPE} probes did not alter the
observable behaviour of CPSCore: all 1,425 test assertions passed
after instrumentation, confirming that the macro introduces no
side effects.
The \texttt{csp\_matcher} tool's reversibility property
(\cref{sec:csp-matcher}) guarantees that instrumentation can be
cleanly removed by applying the corresponding \texttt{remove:}
rule, leaving the source tree byte-identical to its
pre-instrumentation state.

\section{Threats to Validity}

\subsection{Researcher Degrees of Freedom}\label{sec:threats-rdof}

The H1 and H2 experiments surfaced several points at which a definitional
choice could have been made to favour a preferred result: ground-truth
circularity, the timing of the \texttt{SCENARIO\_PRIMARY} scope definition,
F1 aggregation, scenario mix, and the H2 noise/reference circularity
(\cref{sec:h1-corrections,sec:h2-corrections}).
Each was identified and corrected before being treated as a final result
rather than after. This is reported explicitly, as a threat to validity in
its own right: the fact that these risks were caught in this project does
not mean an equivalent hybrid-reconstruction study run by a different team,
or under time pressure, would catch them too.

\subsection{Internal Validity}

\begin{description}
  \item[Reference diagram subjectivity.]
    The manually authored reference diagrams reflect the understanding
    of a single engineer, per scenario.
    An alternative reference produced by a different reviewer might
    differ in the granularity of message names or the selection of
    lifelines.

  \item[Instrumentation coverage.]
    Instrumentation targets (both full and guided) were selected by
    inspecting the static graph, not by running coverage analysis.
    A different selection of entry points would produce a different
    trace, and thus a different H1/H2 result.

  \item[Synthetic timestamps.]
    Timestamps in the static-derived sequence table are synthetic
    (1\,ms spacing from an arbitrary base).
    They encode order correctly for the purposes of sequence diagram
    generation but do not reflect real wall-clock behaviour.
\end{description}

\subsection{External Validity}

\begin{description}
  \item[Single \emph{controlled} system for H1/H2/H3.]
    CPSCore is a moderate-scale framework with a clean modular
    structure.
    The practitioner session on a second, much larger industrial codebase
    (\cref{sec:practitioner}) provides encouraging qualitative evidence
    that the underlying static-extraction and query approach scales and
    transfers. However, it is an unblinded, four-participant interview
    study on a real defect, not a repeat of the controlled H1/H2/H3
    comparisons.
    Applying the full, pre-registered hypothesis tests to a second system
    remains the strongest way to strengthen external validity.

  \item[Renaissance dependency.]
    The current static extraction stage depends on the Renaissance
    semantic analyser, which is an internal TNO-ESI tool.
    Replicating the pipeline in a different organisation would require
    substituting an equivalent \Cpp\ semantic extractor
    (\eg a Clang-based tool) that emits a compatible graph schema.
    The agent skills are schema-agnostic and would require only the
    \texttt{get-neo4j-schema} skill to be re-run against the new schema.

  \item[S4 is constructed, not found, and is H1's sole fault-path/dual-gap
    scenario.]
    The strict-additivity result (fusion strictly exceeds both single sources at once,
    \cref{sec:h1-c4c3}) rests on a single scenario, S4, whose production
    code (a \texttt{boost::signals2} publish/subscribe bridge) was written
    specifically for this experiment rather than discovered in CPSCore's
    existing codebase. It is reported and stratified separately for exactly
    this reason. A larger and more diverse set of fault-path/dual-gap
    scenarios would give more confidence that the underlying mechanism,
    rather than some property of S4 specifically, drives the effect. The
    finding demonstrates that the mechanism is real and
    reproducible under controlled conditions, and corroborates an
    independent, naturally-occurring observation of the same mechanism in
    the practitioner session (\cref{sec:practitioner}), but a larger sample
    of naturally-occurring cross-module-indirection scenarios, ideally from
    a codebase where such indirection is pervasive rather than absent,
    would be needed to estimate how common the strict-additivity effect is
    in practice.
\end{description}

\subsection{Construct Validity}

\begin{description}
  \item[Edge-level evaluation metric.]
    Precision and recall are measured over unique
    $\langle\text{sender},\,\text{receiver}\rangle$ pairs.
    This abstracts over message count and temporal ordering, which
    may underestimate the practical importance of ordering accuracy
    in a sequence diagram.
    A metric that accounts for sequence alignment would provide a
    more complete picture and is identified as future work.

  \item[Noise/coverage circularity in H2.]
    As discussed in \cref{sec:h2-corrections}, the noise metric shares a
    scope definition with the guided-instrumentation slicing criterion,
    which inflates the apparent noise reduction for scenarios where the
    slice and the reference were defined from the same pass. This is
    mitigated, but not eliminated, by treating S3 as an independently
    defined falsification case.
\end{description}

