\chapter{Introduction}\label{ch:introduction}

\section{Context}

Large-scale software-intensive systems developed and maintained by industrial
research organisations are often mission-critical, long-lived, and subject to
continuous evolution \cite{systa:reverse}.
These systems frequently operate in safety-critical, high-reliability, or
performance-sensitive domains such as medical imaging and industrial automation.
In these domains, an incorrect change or a poorly informed architectural decision
can have significant technical and economic consequences.
Consequently, a deep and accurate understanding of system behaviour is essential
before performing refactoring, modernisation, or architectural restructuring.

This understanding is supposed to be captured by the \emph{Interaction Model}: a
formal, scenario-specific description of the message exchanges between components.
Interaction Models are most commonly expressed as sequence diagrams in \ac{UML} or
(more recently) in \sysml.
In practice, Interaction Models are rarely kept up to date with the source code
they describe.
Over time, such systems accumulate complexity due to evolving requirements,
heterogeneous technologies, and extensive use of concurrency and asynchrony
\cite{systa:reverse}.
Design documentation drifts out of sync with the codebase, and diagrams that were
accurate when originally drawn diverge from reality as the system changes.
As a result, engineers are required to reconstruct system behaviour directly from
source code and limited runtime observations. This process is time-consuming,
error-prone, and highly dependent on individual expertise \cite{systa:reverse}.

This challenge is particularly pronounced in event-driven systems.
Asynchronous message passing, callbacks, and indirect control flow obscure
the relationships between interacting components \cite{cornelissen:systematic}.
Understanding which components interact, in what order, and under which conditions
is non-trivial. Yet it is crucial for identifying architectural bottlenecks,
isolating faulty behaviour, and assessing the impact of proposed changes.
From industry's perspective, there is a strong need for accurate and
scenario-specific interaction models that reflect how systems behave under
concrete execution scenarios \cite{egyed:scenario}.
Such models can assist engineers in taking informed architectural and refactoring
decisions, isolating relevant subsystems for targeted improvement, increasing
test coverage for critical interaction paths, and improving long-term
maintainability and knowledge transfer in complex industrial systems.

This project surfaced a concrete illustration of this class of problem: a
real, previously-resolved defect on a large industrial \Cpp\ system
(\cref{sec:scenario-exercise}). A stale proxy object was never freed,
because a reset call was missing from the connection-handling code. As a
result, the system silently stopped delivering change-notification events
after a service restart.
The fault was visible only by following an implicit chain of events that no
existing interaction model showed. Diagnosing it meant manually
reconstructing exactly the kind of model this thesis automates.

The \ac{GDI} project at \ac{TNO}-\ac{ESI} aims to address this problem by using
graph-based program analysis to support the automatic extraction and maintenance of
software architecture models.
This thesis contributes to that effort by developing a hybrid approach that combines
two complementary sources of evidence: the static call graph extracted from source code
and timestamped execution traces captured at runtime.

\section{Problem Statement}

Two complementary approaches are commonly used to analyse software behaviour:
static code analysis and dynamic analysis.
\emph{Static analysis} provides a complete view of potential interactions by
revealing call dependencies and structural relationships. However, it typically
over-approximates behaviour and lacks information about execution order and
runtime conditions \cite{systa:reverse}.
It cannot distinguish between interactions that occur in a specific scenario
and those that occur in every other scenario.
\emph{Trace-based analysis} captures concrete runtime behaviour and is therefore
scenario-specific by construction. However, it is inherently incomplete
\cite{cornelissen:systematic}.
In industrial systems, traces are often sparse or noisy due to limited logging,
performance constraints on instrumentation, or execution paths that were not
exercised during recording.

Although hybrid static-dynamic approaches have been proposed \cite{labiche:hybrid,
cornelissen:systematic}, existing methods primarily use static structure to
\emph{organise} or contextualise an already-observed trace (\cref{ch:related}).
They do not actively infer interactions that the trace never captured.
No existing approach systematically \emph{combines} the two directions at once:
using static structure to recover the interactions a trace missed, and using the
trace to trim an over-approximate static graph down to only what is relevant to
the scenario at hand.
Closing that gap in both directions is the core problem this thesis addresses.

A further complication arises from the publish-subscribe indirection common
in event-driven \Cpp\ systems.
CPSCore, the open-source framework used as the primary case study, routes
inter-component communication through an \ac{IDC}
abstraction layer backed by \texttt{boost::signals2} callbacks and Redis
channels.
A static call graph built from function call sites alone contains edges
\emph{into} the IDC layer but does not reveal which component receives each
published packet. The subscription topology is implicit in callback
registrations rather than explicit in call sites.
Recovering that missing structure is part of the contribution of this thesis
(\cref{sec:macro-resolution}).

\section{Research Questions}

This thesis is guided by the following research questions:

\begin{description}
  \item[\rqone] What intermediate property graph representation can integrate
    static call graphs and timestamped runtime events for interaction model
    reconstruction?

  \item[\rqtwo] How does combining static analysis and incomplete traces
    improve accuracy and completeness of scenario-specific interaction model
    reconstruction?

  \item[\rqthree] How can dependence-based slicing filter trace events and
    constrain static interactions to only scenario-relevant behaviour?
\end{description}

Each research question is paired with a corresponding hypothesis that makes it
testable, and each hypothesis is answered by a dedicated experiment
(\cref{ch:experiments,ch:analysis}):

\begin{description}
  \item[H1 (\rqtwo).] Combining static analysis and dynamic instrumentation
    improves reconstruction accuracy over either source alone.
  \item[H2 (\rqthree).] Code-graph-guided instrumentation improves scenario
    focus by suppressing irrelevant execution detail relative to full,
    unguided instrumentation.
  \item[H3.] Reconstructed interaction models improve agent task performance
    on scenario comprehension, relative to reasoning from raw source code
    alone.
\end{description}

\rqone\ is answered by the design of the property graph schema
(\cref{ch:methods}). \rqtwo\ and \rqthree\ are answered empirically by the H1
and H2 experiments on the CPSCore case study, with H3 assessing whether the
reconstructed models translate into a measurable downstream benefit
(\cref{ch:experiments,ch:analysis}).

\section{Approach}

The proposed approach consists of four stages:

\begin{enumerate}
  \item \textbf{Static extraction.} A call graph is extracted from the \Cpp\
    source code of the target system and stored in a \neo \acl{GDB} using the
    schema described in \cref{sec:pg-schema}.
    For publish-subscribe systems that route messages through callback
    registration (such as CPSCore's IDC layer), this stage additionally
    applies topology resolution to recover the implicit routing structure
    (\cref{sec:macro-resolution}).

  \item \textbf{Trace ingestion.} Timestamped execution traces produced during
    a scenario recording session are parsed and aligned to the entities in the
    static graph, producing a set of observed interactions annotated with
    temporal ordering information.

  \item \textbf{Graph fusion and slicing.} The static and trace graphs are merged
    into a unified \acl{PG}. Dependence-based slicing is applied to the merged
    graph to retain only those nodes and edges that are reachable from the
    scenario entry point, filtering both statically-derived and trace-derived
    interactions that are not causally connected to the scenario under analysis.

  \item \textbf{SysML v2 generation.} The sliced graph is projected into
    \sysml sequence diagrams using a template-driven generator.
    The generated diagrams are written to a text file following the
    \sysml textual notation.
\end{enumerate}

The approach is developed and validated on \emph{CPSCore}, an open-source
event-driven \Cpp\ framework, across four scenarios spanning
happy-path and constructed indirection-boundary behaviour
(\cref{ch:experiments,ch:analysis}).

\section{Outline}

The remainder of this thesis is structured as follows.
\Cref{ch:background} introduces the technical concepts underpinning the approach:
\sysml Interaction Models, property graphs, static call-graph analysis, execution
tracing, and dependence-based slicing.
\Cref{ch:related} surveys existing work on program comprehension, architecture
recovery, and trace-based diagram synthesis. It positions this thesis within
that landscape.
\Cref{ch:methods} describes the design of the approach in detail, including the
property graph schema, the fusion algorithm, and the slicing procedure.
\Cref{ch:experiments} presents the implementation and the results of applying
the approach to the CPSCore case study.
\Cref{ch:analysis} analyses those results with respect to the three research
questions and the three hypotheses. It discusses threats to validity and
reflects on the limitations of the approach.
\Cref{ch:conclusion} summarises the contributions, discusses future work, and
concludes the thesis.
